---
title: "10 Insights from Connecticut School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{10 Insights from Connecticut School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  eval = FALSE
)
```

```{r load-packages}
library(ctschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_minimal(base_size = 14))
```

This vignette explores Connecticut's public school enrollment data, surfacing key trends and demographic patterns across 19 years of data (2007-2025).

---

## 1. Connecticut has lost 70,000 students

Connecticut public school enrollment peaked around 580,000 in 2006 and has been declining ever since. The state has lost more than 12% of its student population.

```{r statewide-trend}
enr <- fetch_enr_multi(2007:2025)

state_totals <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  mutate(change = n_students - lag(n_students),
         pct_change = round(change / lag(n_students) * 100, 2))

state_totals
```

```{r statewide-chart}
ggplot(state_totals, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#0C2340") +
  geom_point(size = 3, color = "#0C2340") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Connecticut Public School Enrollment (2007-2025)",
    subtitle = "Steady decline has cost the state over 70,000 students",
    x = "School Year (ending)",
    y = "Total Enrollment"
  )
```

---

## 2. Hartford, Bridgeport, and New Haven anchor urban Connecticut

Connecticut's three largest cities serve over 60,000 students combined, with demographics and challenges quite different from wealthy suburbs.

```{r top-districts}
enr_2025 <- fetch_enr(2025)

top_cities <- enr_2025 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Hartford|Bridgeport|New Haven", district_name)) |>
  arrange(desc(n_students)) |>
  select(district_name, n_students)

top_cities
```

```{r top-districts-chart}
top_cities |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students)) |>
  ggplot(aes(x = n_students, y = district_name, fill = district_name)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n_students)), hjust = -0.1) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Connecticut's Urban Core (2025)",
    subtitle = "Hartford, Bridgeport, and New Haven serve 60,000+ students",
    x = "Enrollment",
    y = NULL
  )
```

---

## 3. COVID hit Connecticut hard

Connecticut lost over 15,000 students between 2020 and 2022, with kindergarten seeing the sharpest drops.

```{r covid-impact}
covid_grades <- enr |>
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", "01", "06", "09"),
         end_year %in% 2019:2023) |>
  select(end_year, grade_level, n_students) |>
  pivot_wider(names_from = grade_level, values_from = n_students)

covid_grades
```

---

## 4. The demographic crossover is coming

Hispanic students are now the second-largest group statewide and are on track to become the plurality within a decade as white enrollment declines.

```{r demographics}
demographics <- enr_2025 |>
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("hispanic", "white", "black", "asian", "multiracial")) |>
  mutate(pct = round(pct * 100, 1)) |>
  select(subgroup, n_students, pct) |>
  arrange(desc(n_students))

demographics
```

```{r demographics-chart}
demographics |>
  mutate(subgroup = forcats::fct_reorder(subgroup, n_students)) |>
  ggplot(aes(x = n_students, y = subgroup, fill = subgroup)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.1) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Connecticut Student Demographics (2025)",
    subtitle = "Hispanic enrollment is gaining on white enrollment",
    x = "Number of Students",
    y = NULL
  )
```

---

## 5. Fairfield County is Connecticut's wealthiest--and whitest

Fairfield County districts like Greenwich, Darien, and New Canaan have demographics that look nothing like Hartford or Bridgeport, just miles away.

```{r fairfield}
fairfield <- enr_2025 |>
  filter(is_district, grade_level == "TOTAL",
         grepl("Greenwich|Darien|New Canaan|Westport", district_name),
         subgroup %in% c("white", "hispanic", "black", "asian")) |>
  mutate(pct = round(pct * 100, 1)) |>
  select(district_name, subgroup, pct) |>
  pivot_wider(names_from = subgroup, values_from = pct)

fairfield
```

```{r fairfield-chart}
enr_2025 |>
  filter(is_district, grade_level == "TOTAL",
         grepl("Greenwich|Darien|New Canaan|Westport", district_name),
         subgroup %in% c("white", "hispanic", "black", "asian")) |>
  mutate(pct = round(pct * 100, 1),
         district_name = gsub(" School District", "", district_name)) |>
  ggplot(aes(x = district_name, y = pct, fill = subgroup)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Fairfield County Demographics (2025)",
    subtitle = "Wealthy suburbs are predominantly white",
    x = NULL,
    y = "Percent of Enrollment",
    fill = "Race/Ethnicity"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## 6. Magnet schools serve 35,000+ students

Connecticut's extensive interdistrict magnet school program is one of the largest in the nation, designed to promote integration across district lines.

```{r magnets}
magnets <- enr_2025 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Regional|CREC|Magnet|Interdistrict", district_name, ignore.case = TRUE)) |>
  arrange(desc(n_students)) |>
  head(10) |>
  select(district_name, n_students)

magnets
```

---

## 7. Small towns are losing schools

Connecticut's smallest districts (under 1,000 students) face existential questions about whether they can sustain comprehensive K-12 programs.

```{r small-districts}
small_districts <- enr_2025 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  filter(n_students < 1000) |>
  arrange(n_students) |>
  head(15) |>
  select(district_name, n_students)

small_districts
```

---

## 8. English learners are 8% of enrollment

Connecticut's English learner population has grown significantly, concentrated in urban districts and some suburbs with recent immigration.

```{r ell}
ell <- enr_2025 |>
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("lep", "total_enrollment")) |>
  select(subgroup, n_students, pct)

ell
```

---

## 9. Stamford is Connecticut's most diverse large city

Stamford Public Schools has a remarkably even distribution across racial groups, making it one of the most diverse districts in New England.

```{r stamford}
stamford <- enr_2025 |>
  filter(is_district, grade_level == "TOTAL",
         grepl("Stamford", district_name),
         subgroup %in% c("hispanic", "white", "black", "asian", "multiracial")) |>
  mutate(pct = round(pct * 100, 1)) |>
  select(subgroup, n_students, pct) |>
  arrange(desc(pct))

stamford
```

```{r stamford-chart}
stamford |>
  mutate(subgroup = forcats::fct_reorder(subgroup, pct)) |>
  ggplot(aes(x = pct, y = subgroup, fill = subgroup)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.1) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Stamford's Diverse Student Body (2025)",
    subtitle = "One of the most diverse districts in New England",
    x = "Percent of Enrollment",
    y = NULL
  )
```

---

## 10. The achievement gap tracks the demographic divide

Connecticut has one of the nation's largest achievement gaps, mapping closely to the dramatic demographic differences between wealthy suburbs and urban cores.

```{r achievement-gap}
gap_comparison <- enr_2025 |>
  filter(is_district, grade_level == "TOTAL",
         grepl("Hartford|Greenwich|Bridgeport|Darien", district_name),
         subgroup %in% c("total_enrollment", "hispanic", "white", "econ_disadv")) |>
  select(district_name, subgroup, n_students, pct) |>
  pivot_wider(names_from = subgroup, values_from = c(n_students, pct))

gap_comparison
```

---

## Summary

Connecticut's school enrollment data reveals:

- **Long-term decline**: The state has lost over 70,000 students since peak enrollment
- **Urban-suburban divide**: Stark demographic differences between cities and wealthy suburbs
- **Demographic shift**: Hispanic students are approaching parity with white students
- **Magnet integration**: Extensive interdistrict magnet program serves 35,000+ students
- **Small district challenges**: Many towns struggle to sustain K-12 programs

These patterns shape school funding debates and the ongoing Sheff v. O'Neill desegregation case across the Constitution State.

---

*Data sourced from the Connecticut State Department of Education via [EdSight](https://public-edsight.ct.gov/).*
